---
title: "Appendix: the nickname policy, and the ablation that decided it"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Appendix: the nickname policy, and the ablation that decided it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mysterynpi)
```

This appendix documents a decision, not a feature. On 2026-09-07 a
frozen-matcher ablation (matcher `fa7216f`, dictionary `2026-09-06.1`)
asked the question the nickname machinery had been engineered around
without answering: **does curated nickname expansion actually improve
identity linkage, and at what false-positive cost?** The answer split
cleanly by layer, and the split *is* the policy:

* **Verdict layer** (`nickname_agreement()`): **retained globally.**
* **Candidate expansion** (`npi_search()`'s `curated_one_hop`):
  **review-only**, gated to source classes that plausibly record go-by
  names, never sufficient for auto-acceptance.

On 2026-09-09 the policy was superseded only to separate initials from
nickname evidence. The 327-person adjudication instrument and dictionary
edges were not changed.

The policy ships as data so any consumer can cite it:

```{r}
str(NICKNAME_POLICY, max.level = 1)
```

## The verdict-layer result, recomputed here

The evidence below is not quoted — it is **recomputed on every build of
this vignette** from `ROSTER_BENCHMARK` (190 pairs, truth by
construction; see `vignette("roster-benchmark")`), using that vignette's
reference policy with only the given-name rule swapped. Condition A
ablates the nickname table (exact full-token equality only; initials are
uninformative); condition B is `nickname_agreement()` as shipped.

```{r}
decisions <- function(given_rule) {
  b  <- ROSTER_BENCHMARK
  ex <- extract_suffix(b$roster_name)
  p  <- parse_person(ex$name)
  axes <- data.frame(
    surname = surname_agreement(p$last, b$npi_last,
                                middle_a = p$middle, middle_b = b$npi_middle),
    given   = given_rule(sub(" .*", "", p$first), b$npi_first),
    middle  = middle_agreement(middle_tokens(p$middle),
                               middle_tokens(b$npi_middle)),
    suffix  = suffix_agreement(ex$suffix, b$npi_suffix),
    gender  = gender_agreement(b$roster_gender, b$npi_gender),
    license = license_agreement(b$roster_license, b$roster_state,
                                b$npi_license, b$npi_state))
  excused <- mapply(function(mt, nl)
    length(intersect(mt, surname_tokens(nl))) > 0,
    middle_tokens(p$middle), b$npi_last)
  conflict <- axes$surname == "conflicts" | axes$given == "conflicts" |
    (axes$middle == "conflicts" & !excused) | axes$suffix == "conflicts"
  name_ok <- axes$surname == "corroborates" & axes$given == "corroborates"
  ifelse(conflict, "reject",
  ifelse(!name_ok, "review",
  ifelse(axes$gender == "conflicts", "review", "accept")))
}
no_table <- function(a, b) {
  norm <- function(x) gsub("[.]", "", name_key(x))
  ka <- norm(a); kb <- norm(b)
  vapply(seq_along(ka), function(i) {
    x <- ka[i]; y <- kb[i]
    if (!has_name_information(x) || !has_name_information(y))
      return("uninformative")
    if (nchar(x) == 1L || nchar(y) == 1L)
      return("uninformative")
    if (x == y) return("corroborates")
    "conflicts"
  }, character(1))
}
A <- decisions(no_table)
B <- decisions(nickname_agreement)
table(A, truth = ROSTER_BENCHMARK$truth)
table(B, truth = ROSTER_BENCHMARK$truth)
```

Thirty-three adjudicated true matches still depend on governed nickname
edges, at zero false accepts. Removing initials from nickname evidence
moves initial/full-name rows to review rather than letting them masquerade
as nickname corroboration. These exact counts are pinned in
`tests/testthat/fixtures/ablation/ablation_pins_v2.csv`, whose checksum
is coupled to the policy id. The original
`tests/testthat/fixtures/ablation/ablation_pins_v1.csv` remains in the
tree as superseded evidence; editing the active evidence without a policy
supersession fails CI.

An honest caveat: this benchmark is constructed, and its nickname
families exist by design. It proves the machinery does what it claims at
zero measured cost; real-world *prevalence* came from the production
crosswalks below.

## The candidate-layer result (external data, summarized)

Candidate generation was measured against 221 human-adjudicated
roster-to-NPI links with a 40,000-row NPPES candidate pool ("adjudicated
by a human, NOT produced by any matcher"; fixtures live outside this
package):

| endpoint | none | curated one-hop | increment |
|---|---:|---:|---:|
| anchors with true NPI in candidates | 163 | 164 | **+1** |
| total candidates | 357 | 435 | **+78** (77 false) |
| incremental PPV | — | 0.013 | — |
| anchors gaining review burden | — | 30/221 | 13.6% |

The one rescue (DAN, recorded as a go-by name on the roster, DANIEL in
the registry) is the exception that defines the rule: the nickname
channel exists only where a source *writes* informal names. A live-registry
replication found 0 rescues and +76% review burden. In production
crosswalks, an official legal-name roster (22,309 certificants) carried
14 parenthesized nicknames and zero nickname-dependent links; a scraped
semi-official roster carried 26 nickname-accepted links in 59,387 —
twelve of them on name evidence alone.

Hence the gate, resolved against one canonical registry — never by
string-matching source names, with `unknown` failing closed:

```{r}
SOURCE_CLASSES
```

```{r, error = TRUE}
npi_search(first_name = "bill", last_name = "smith",
           name_expansion = "curated_one_hop")   # formal_record: refused
```

## What a permitted expansion carries

With `source_class = "informal_capable"`, every returned row is
attributable: the plan itself shows each query and the exact dictionary
edge that licensed it —

```{r}
nickname_variants("BILL")[1:4, ]
```

— and result rows carry `found_by_queries` / `found_by_edges` lineage
(the identity path written as the literal `input`), plus
`acceptance_contribution`: `none`, `supporting`, or `nickname_only`. The
contract with the acceptance layer is `assert_nickname_policy()`: a
`nickname_only` candidate, or any expansion-influenced row with broken
lineage, can never be auto-accepted — it goes to a human.

## Negative controls

All ten known-ghost fixtures stay rejected and unreachable, recomputed
here:

```{r}
ghosts <- data.frame(
  a = c("MARVIN", "GEORGE", "CHRISTINA", "PATRICIA", "DANIELLE",
        "ROBERT", "HAROLD", "ALBERT", "ELISABETH", "JANE"),
  b = c("MORRIS", "GRETA", "CHRISTOPHER", "PATRICK", "DANIEL",
        "WILLIAM", "HENRY", "ALEXANDER", "ELIZABETH", "JOAN"))
data.frame(ghosts,
  verdict = nickname_agreement(ghosts$a, ghosts$b),
  reachable = mapply(function(x, y)
    y %in% nickname_variants(x)$queried_first_name, ghosts$a, ghosts$b))
```

## Governed edges: the ROBERT>BILL record

`ROBERT>BILL` is the corpus's most-debated edge. Measured everywhere it
could appear: zero rescues; its only observed act is manufacturing
ROBERT candidates in BILL searches. It stays — as a governed, versioned
data decision, its downside bounded by the review-only rule, its record
carried in the policy object:

```{r}
NICKNAME_POLICY$governed_edges
```

## Provenance

The complete study — design pre-registration, four data sources, the
per-edge ledger (34 edges fired, one rescue), source-pair matrix,
limitations — is archived with the evidence
(`tools/ablation/README.md` points at everything). The verdict-layer
harness ships in `tools/ablation/exp2_verdict_layer.R` and reruns
anywhere this package installs. Matching was frozen at `fa7216f`
throughout; nothing in the matcher, thresholds, or dictionary changed
during the study.
