---
title: "The roster benchmark: 190 labeled pairs, and what they measure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The roster benchmark: 190 labeled pairs, and what they measure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mysterynpi)
```

No public benchmark exists for physician-roster-to-NPPES name matching:
NPPES describes real providers, so a labeled gold standard would name real
people, and nobody ships one. `ROSTER_BENCHMARK` is the synthetic
alternative -- 190 (roster record, registry candidate) pairs, truth assigned
**by construction** in reviewable code, each pair carrying the name of the
defect family it encodes. The authoring rule: *a pair whose truth a reviewer
could not assign did not go in.* (That rule has teeth. The first draft's
spelling-trap family paired `Julia Bennett` against `JULIE BENNETT` with
nothing else recorded and called it a nonmatch -- but no reviewer could know
that, and the nickname corpus records `JULIA`/`JULIE` as substitutable
usage. The shipped rows carry conflicting middle initials, so the deciding
evidence is real and spelling alone decides nothing, in either direction.)

```{r}
table(ROSTER_BENCHMARK$family, ROSTER_BENCHMARK$truth)
```

## A reference policy from the package's own rules

Pairs are pre-formed (blocking is the pipeline's job, not this package's),
so the policy is pure agreement: parse the roster's free text, compare every
axis, veto on hard conflicts, quarantine gender conflicts at the top, accept
what the name evidence earns.

```{r}
b  <- ROSTER_BENCHMARK
ex <- extract_suffix(b$roster_name)          # BEFORE the noise strip eats it
p  <- parse_person(ex$name)

axes <- data.frame(
  surname = surname_agreement(p$last, b$npi_last,
                              middle_a = p$middle, middle_b = b$npi_middle),
  given   = nickname_agreement(sub(" .*", "", p$first), b$npi_first),
  middle  = middle_agreement(middle_tokens(p$middle),
                             middle_tokens(b$npi_middle)),
  suffix  = suffix_agreement(ex$suffix, b$npi_suffix),
  gender  = gender_agreement(b$roster_gender, b$npi_gender),
  license = license_agreement(b$roster_license, b$roster_state,
                              b$npi_license, b$npi_state))
```

One refinement the maiden-as-middle family forces, and which any real
pipeline needs: when the roster's middle token **is** the registry's
surname, it is a former surname sitting in a middle slot -- the surname
rescue has already used it -- and its middle-name "conflict" must not veto:

```{r}
excused <- mapply(function(mt, nl) length(intersect(mt, surname_tokens(nl))) > 0,
                  middle_tokens(p$middle), b$npi_last)

conflict <- axes$surname == "conflicts" |
  axes$given == "conflicts" |
  (axes$middle == "conflicts" & !excused) |
  axes$suffix == "conflicts"
name_ok <- axes$surname == "corroborates" & axes$given == "corroborates"

decision <- ifelse(conflict, "reject",
            ifelse(!name_ok, "review",
            ifelse(axes$gender == "conflicts", "review", "accept")))
table(decision, truth = b$truth)
```

Every accept is a true match, every reject a true nonmatch, and the review
queue is exactly the stale-gender family: true matches whose recorded
gender codes conflict, routed to a person instead of deleted -- the policy
`vignette("vetoes-and-quarantine")` argues for, doing its job.

Two honest caveats. First, this is the package evaluated on a corpus built
from the package's own defect taxonomy -- the families are the failure
modes these rules were written against, so clean separation is
*consistency*, not proof of generality; foreign corpora like
[WINKLER_CENSUS] exist to be harder. Second, the benchmark measures the
agreement rules, not blocking: every pair here was already formed.

The construction itself paid once already: the first evaluation run
rejected three formatting-family matches because `extract_suffix()` was
eating the comma that `parse_person()`'s "Last, First" reversal needs.
That defect is now fixed and pinned in the suffix tests -- which is the
benchmark working as designed.

## From decisions to publishable numbers

The clerical machinery turns the accepted and quarantined pairs into
per-class precision with exact intervals. Class 1: license corroborates;
class 2: middle corroborates; class 3: name evidence alone; class 4:
quarantined for review.

```{r}
kept <- data.frame(
  id = b$pair_id, candidate = paste0("N", seq_len(nrow(b))),
  evidence_class = ifelse(decision == "review", 4L,
                   ifelse(axes$license == "corroborates", 1L,
                   ifelse(axes$middle == "corroborates", 2L, 3L))),
  roster_name = b$roster_name,
  stringsAsFactors = FALSE)[decision != "reject", ]

s <- clerical_sample(kept, n_per_class = 12, seed = 20260905)
verdicts <- data.frame(
  review_id = s$key$review_id,
  is_match = b$truth[match(s$key$id, b$pair_id)] == "match")
clerical_precision(s$key, verdicts)
```

In a real study the `is_match` column comes from a blinded human reviewer,
not from stored truth; here the stored truth stands in to show the
workflow. The class-4 row is the payoff to notice: quarantine turned six
would-be deletions into six reviewed true matches.

## The harder corpus, and the honest number

`WINKLER_CENSUS` measures a failure mode the roster benchmark deliberately
does not: TYPOS. Winkler built his synthetic census pairs by injecting
spelling corruption -- `BELL`/`VELL`, `SCCHWARTZ`/`SCHWARTZ`,
`CHARLS`/`CHARLES` -- which is precisely the corruption class this
package's rules refuse to bridge, because every edit-distance tolerance
they ever tried admitted genuinely different people. So run the same
policy over his 327 labeled matches and 582 same-household hard negatives,
and publish what happens:

```{r}
w <- WINKLER_CENSUS
a <- w[w$relation == "A", ]; a <- a[!duplicated(a$id), ]
b <- w[w$relation == "B", ]; b <- b[!duplicated(b$id), ]
m <- merge(a, b, by = "id", suffixes = c("_a", "_b"))

verdicts <- function(d) {
  ax <- data.frame(
    surname = surname_agreement(d$surname_a, d$surname_b),
    given   = nickname_agreement(d$given_a, d$given_b),
    middle  = middle_agreement(middle_tokens(d$middle_a),
                               middle_tokens(d$middle_b)))
  conflict <- ax$surname == "conflicts" | ax$given == "conflicts" |
    ax$middle == "conflicts"
  name_ok <- ax$surname == "corroborates" & ax$given == "corroborates"
  ifelse(conflict, "reject", ifelse(name_ok, "accept", "review"))
}
table(true_pairs = verdicts(m))

hn <- merge(a, b, by = c("house", "street"), suffixes = c("_a", "_b"))
hn <- hn[hn$id_a != hn$id_b, ]
table(household_negatives = verdicts(hn))
```

Recall `r sprintf("%.1f%%", 100 * mean(verdicts(m) == "accept"))` on true
pairs; zero false accepts on the household negatives. That is not a bug to
fix -- it is the trade this package makes, stated as a measurement instead
of a shrug. Typo-corrupted true matches are recovered, in the pipelines
this package serves, at the BLOCKING layer (a fuzzy surname pass that
*generates* a candidate ranked below exact evidence) and by non-name
fields -- license, address, birth facts -- never by loosening what a name
agreement means. Note what did NOT fail: `TRICIA`/`PATRICIA` is a recorded
nickname edge and corroborated; that pair died on its `SCCHWARTZ` surname
typo, which no nickname table should rescue.

The benchmark ships as plain CSV for use outside R:

```{r, eval = FALSE}
system.file("extdata", "roster_benchmark.csv", package = "mysterynpi")
```
