---
title: "Vetoes and quarantine: using conflicts without deleting people"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vetoes and quarantine: using conflicts without deleting people}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mysterynpi)
```

The agreement rules in this package report **verdicts** -- `"corroborates"`,
`"conflicts"`, `"uninformative"` -- and deliberately do not decide what a
verdict does. That split matters most for the two rules whose conflicts are
built on fields that are *sometimes wrong about true matches*:

* **Gender.** A recorded code can be stale or simply wrong: data entry, a
  transition the registry has not caught up with, a source that coded the
  practice owner rather than the provider. The linkage literature's guidance
  is consistent: disagreement on one error-prone field should not be
  unrecoverable when everything else agrees at the strongest class.
* **Surnames.** Marriage moves a surname wholesale. In a cohort where that is
  plausible -- most physician cohorts -- total surname disagreement alongside
  top-class agreement on given name and middle name is a case for review,
  not deletion. (`surname_agreement()`'s maiden-as-middle rescue catches the
  cases where the old surname survives in a middle slot; this vignette is
  about the ones it cannot.)

The policy this vignette recommends: **a conflict vetoes in the weaker
evidence classes, and quarantines at the strongest one.** Fully deterministic,
and every action it takes is recorded.

## A cohort of two people

`r1`'s only strong candidate carries a conflicting gender code. `r2` is the
clean case.

```{r}
cands <- data.frame(
  id              = c("r1",   "r1",   "r2",   "r2"),
  candidate       = c("N100", "N101", "N200", "N201"),
  gender_roster   = c("F",    "F",    "F",    "F"),
  gender_registry = c("M",    "F",    "F",    "M"),
  evidence_class  = c(1L,     2L,     1L,     2L),
  stringsAsFactors = FALSE)
cands$gender <- gender_agreement(cands$gender_roster, cands$gender_registry)
cands
```

## The policy that looks safe and is not

A hard veto drops every `"conflicts"` row before resolution:

```{r}
hard <- cands[cands$gender != "conflicts", ]
resolve_ordered_classes(hard, tiebreak = "candidate")$resolved[
  , c("id", "candidate", "evidence_class")]
```

`r1` resolves -- to `N101`, its *class-2* candidate. If the registry's code
for `N100` was the error (a true match with a wrong byte), the hard veto did
not just lose a match: it silently published a **different identity** for a
real person, with nothing in the output recording that a stronger candidate
was deleted on one field.

## Veto below, quarantine at the top

```{r}
conflict  <- cands$gender == "conflicts"
top_class <- cands$evidence_class == min(cands$evidence_class)

vetoed      <- cands[conflict & !top_class, ]        # recorded, then removed
quarantined <- unique(cands$id[conflict & top_class])

kept <- cands[!(conflict & !top_class) & !cands$id %in% quarantined, ]
res  <- resolve_ordered_classes(kept, tiebreak = "candidate")

res$resolved[, c("id", "candidate", "evidence_class")]
quarantined
vetoed[, c("id", "candidate", "evidence_class")]
```

`r2` still resolves at class 1, and its class-2 rival's veto is **recorded**
in `vetoed` -- the [count_rivals()] discipline: a row published as "no
candidate" must be distinguishable from a row whose candidate was deleted on
one field. `r1` resolves to nothing; it goes to review carrying the reason.

## Quarantine feeds review, review feeds numbers

The quarantined pairs are exactly what [clerical_sample()] is for: a blinded
reviewer decides whether `N100` is `r1` with a wrong code or a different
person, and [clerical_precision()] turns those decisions into the per-class
precision a methods section can print. The conflict was never ignored and
never fatal -- it changed *who decides*, from the pipeline to a person.
